Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-03-08 paper/paper1/ms.tex
Make the performance a subsection, not section
Replace placeholder text with actual links and citations

A key new enabling technology breaks these classical tradeoffs in data-model comparisons for astronomical spectroscopy.  Automatic differentiation \citep[``autodiff'',][]{2015arXiv150205767G} and its affiliated backpropagation algorithm has revolutionized machine learning and neural network architecture design, and is increasingly applied in astrophysical data analysis contexts, \emph{e.g.} kernel phase coronography with \texttt{poppy} \citep{2021ApJ...907...40P}, and exoplanet orbit fitting with \texttt{exoplanet} \citep{2021JOSS....6.3285F}.  Of the spectroscopy frameworks mentioned above, the \texttt{TensorFlow}-based \citep{tensorflow2015-whitepaper} \texttt{wobble} and the \texttt{JAX}-based \censorbox{jax2018github} \texttt{ExoJAX} employ autodiff technology.  \texttt{wobble} treats each pixel as a tunable control point, producing $\sim10^5$ parameters for a modern stellar spectrum.  The \texttt{ExoJAX} framework has only $\sim$dozens of tunable parameters that describe the fundamental physical properties controlling a brown dwarf atmosphere.  These two autodiff-aware frameworks span the extreme ends of non-parametric and parametric modeling for spectroscopy.

In this paper we show that autodiff-aware semi-empirical models offer an appealing middle ground: informed from self-consistent models, but refined with data.  In Section \ref{methodology} we introduce the \emph{blas\'e} forward-model design and its \texttt{PyTorch}-based \cite{2019arXiv191201703P} implementation.  We show how to supplant the computationally expensive radiative transfer by cloning pre-computed synthetic spectral models.  In Section \censorbox{XX} we describe how to fine-adjust the cloned models to data in a transfer-learning step.  Section \censorbox{XX} exhibits some injection-recovery type tests on simulated data of known properties.  In Section \censorbox{XX} we show real-world data from high-grasp near-IR spectrographs.  Finally we discuss many conceivable extensions to the framework, and directions for future work in Section \censorbox{XX}.

@@

\subsection{Voigt profile evaluation}
The Voigt profile $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ can be computed in exact closed-form using the Fadeeva function\authorcomment1{citation}. Evaluation of the Fadeeva function can be computationally costly, and so approximate forms may be desirable. Here we adopt the pseudo-Voigt approximation described in \censor{some paper}\authorcomment1{look up reference to PseudoVoigt}.

@@

\section{Semi-empirical models with transfer learning techniques}

@@

where $\zeta$ is the convolution kernel for rigid body rotation \censorbox{cite ExoJAX} and $\bm{v}$ is the spectral axis represented as relative velocity coordinates.

@@

The instrumental resolving power, $R$, acts as a convolution with a Gaussian line profile of width $\sigma=$\censorbox{todo}.  Real astronomical instruments usually have wavelength-dependent resolving power, which complicates the implementation.  The extent to which this effect matters will depend on the science case.  For now, \emph{blas\'e} simply assumes a fixed resolving power.

@@

\section{Performance of transfer learning}
\begin{outline}
    \1 What is the residual level with bare PHOENIX (i.e. tuning continuum only, not lines)?
    \1 What is the residual level with no regularization (i.e. tune all lines and continuum)?
    \2 Should be near-zero except for missing lines
    \2 What should we do about missing lines?
    \1 What is the residual level with modest regularization? What is the typical change to the line?
    \2 Plot of cloned FWHM versus FWHM transferred
    \2 The pseudo-Voigt line properties can be analytically integrated to give an equivalent width
    \2 Plot of cloned EW (before) versus transfer EW (after)
\end{outline}

\section{Discussion}

@@

\begin{outline}
    \1 Is PseudoVoigt Adequate?
    \1 What performance level is adequate? (depends on your application)
    \1 What about PRV applications
    \1 What about the choice of minibatch size?
    \1 How would a sparse tensor change the computational performance?
    \1 Are we using the GPU effectively? Are we memory bandwidth limited?
    \1 Should we update the line center positions?
\end{outline}

\subsection{Limitations}
\begin{outline}
    \1 What to do when continuum is absent even in at the native resolution of precomputed synthetic model?
    \1 Relevant to brown dwarfs and possibly M dwarfs molecular bands
    \1 Bandwidth limitations
    \1 Interplay with tellurics
\end{outline}

>>>>

A key new enabling technology breaks these classical tradeoffs in data-model comparisons for astronomical spectroscopy.  Automatic differentiation \citep[``autodiff'',][]{2015arXiv150205767G} and its affiliated backpropagation algorithm has revolutionized machine learning and neural network architecture design, and is increasingly applied in astrophysical data analysis contexts, \emph{e.g.} kernel phase coronography with \texttt{poppy} \citep{2021ApJ...907...40P}, and exoplanet orbit fitting with \texttt{exoplanet} \citep{2021JOSS....6.3285F}.  Of the spectroscopy frameworks mentioned above, the \texttt{TensorFlow}-based \citep{tensorflow2015-whitepaper} \texttt{wobble} and the \texttt{JAX}-based \citep{jax2018github} \texttt{ExoJAX} employ autodiff technology.  \texttt{wobble} treats each pixel as a tunable control point, producing $\sim10^5$ parameters for a modern stellar spectrum.  The \texttt{ExoJAX} framework has only $\sim$dozens of tunable parameters that describe the fundamental physical properties controlling a brown dwarf atmosphere.  These two autodiff-aware frameworks span the extreme ends of non-parametric and parametric modeling for spectroscopy.

In this paper we show that autodiff-aware semi-empirical models offer an appealing middle ground: informed from self-consistent models, but refined with data.  In Section \ref{methodology} we introduce the \emph{blas\'e} forward-model design and its \texttt{PyTorch}-based \cite{2019arXiv191201703P} implementation.  We show how to supplant the computationally expensive radiative transfer by cloning pre-computed synthetic spectral models.  In Section \ref{transferLearn} we describe how to fine-adjust the cloned models to real-world data from high-grasp near-IR spectrographs, using a transfer-learning step.  Section \ref{secInjRec} exhibits injection-recovery tests on simulated data of known properties, revealing the scalability and fundamental limits of the technique.  Finally we discuss many conceivable extensions to the framework, and directions for future work in Section \ref{secDiscuss}.

@@

\subsection{Voigt profile evaluation}
The Voigt profile $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ can be computed in exact closed-form using the Fadeeva function\authorcomment1{citation}. Evaluation of the Fadeeva function can be computationally costly, and so approximate forms may be desirable. Here we adopt the pseudo-Voigt approximation \citep{Ida:nt0146}.

@@

\section{Semi-empirical models with transfer learning techniques}\label{transferLearn}

@@

where $\zeta$ is the convolution kernel for rigid body rotation \citep{2022ApJS..258...31K} and $\bm{v}$ is the spectral axis represented as relative velocity coordinates.

@@

The instrumental resolving power, $R$, acts as a convolution with a Gaussian line profile of width $\sigma=\frac{c}{2.355 R}$.  Real astronomical instruments usually have wavelength-dependent resolving power, which complicates the implementation.  The extent to which this effect matters will depend on the science case.  For now, \emph{blas\'e} simply assumes a fixed resolving power.

@@

\subsection{Performance of transfer learning}

\begin{outline}
    \1 What is the residual level with bare PHOENIX (i.e. tuning continuum only, not lines)?
    \1 What is the residual level with no regularization (i.e. tune all lines and continuum)?
    \2 Should be near-zero except for missing lines
    \2 What should we do about missing lines?
    \1 What is the residual level with modest regularization? What is the typical change to the line?
    \2 Plot of cloned FWHM versus FWHM transferred
    \2 The pseudo-Voigt line properties can be analytically integrated to give an equivalent width
    \2 Plot of cloned EW (before) versus transfer EW (after)
\end{outline}

\section{Injection/recovery tests}\label{secInjRec}
Here we conduct injection/recovery tests.

\section{Discussion}\label{secDiscuss}

@@

\begin{outline}
    \1 Is PseudoVoigt Adequate?
    \1 What performance level is adequate? (depends on your application)
    \1 What about PRV applications
    \1 Are we using the GPU effectively? Are we memory bandwidth limited?
    \1 Should we update the line center positions?
\end{outline}

\subsection{Limitations}
\begin{outline}
    \1 What to do when continuum is absent even at the native resolution of precomputed synthetic model?
    \1 Relevant to brown dwarfs and possibly M dwarfs molecular bands
    \1 Bandwidth limitations
    \1 Interplay with tellurics
\end{outline}
